---
title: "`lme4u`: An R Package"
subtitle: "Interpretation and Diagnostics for `lme4`"
author: "Holly Cui"
format: revealjs
editor: visual
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(lme4u)
library(lme4)
library(tidyverse)
```

# Introduction

## Mixed-Effects Modeling

Analyze data where observations are structured within groups, with fixed and random effects. 

<br>

```{r, eval=FALSE, echo=TRUE, class = "fragment"}
lm(Score ~ Instruction, data = mydata)
```

<br>

```{r, eval=FALSE, echo=TRUE, class = "fragment"}
lme4::lmer(Score ~ Instruction + (1 | School), data = mydata)
```

<br>

```{r, eval=FALSE, echo=TRUE, class = "fragment"}
lme4::glmer(Score ~ Instruction + (1 | School), family = binomial(link = "logit"), data = mydata)
```

## Why `lme4u`?

- **Interprets mixed-effects models**: provides user-friendly interpretations for fixed and random effects.
- **Assumption diagnostics**: automates residual analysis and visualizations for heterogeneity checks.
- **Hypothesis Testing**: facilitates hypothesis testing specific to mixed structure.
- **Example Data**: offers a cleaned dataset curated for the purpose of mixed-effect modeling.

# Data

A brand new dataset for mixed-effect modeling

## `star`

A cleaned subset of the Project [STAR](https://search.r-project.org/CRAN/refmans/AER/html/STAR.html) (student-teacher achievement ratio) dataset.

```{r, echo=TRUE}
glimpse(star)
```

# Objective and Modeling

## Question we ask ...

::: {style="font-size: 70%;"}
> _Does class size impact students' math test performance, and how does this effect vary across schools?_

:::

$$
\scriptsize \text{MathScore}_{ij} = \beta_0 + \beta_1(\text{ClassType}_{ij}) + b_{0j} + b_{1j}(\text{ClassType}_{ij}) + \epsilon_{ij}
$$

- $\tiny i \text{ indexes students},\ j \text{ indexes schools.}$
- $\tiny \beta_0, \beta_1 \text{ are fixed intercept and slope.}$
- $\tiny b_{0j} \sim N(0, \tau^2_{b0}) \text{ is a random intercept for each school.}$
- $\tiny b_{1j} \sim N(0, \tau^2_{b1}) \text{ is a random slope for ClassType in each school.}$
- $\tiny \epsilon_{ij} \sim N(0, \sigma^2_{\epsilon}) \text{ is the residual error (within-school heterogeneity).}$

## Modeling

In R using `lme4`, this would be implemented as:

```{r, echo=TRUE}
lme4::lmer(math ~ cltype + (cltype | school_id), data = star)
```

# Package Features

<br>

```{r, eval=FALSE, echo=TRUE}
# Installation 
devtools::install_github("hollyyfc/lme4u")
library(lme4u)
```

 
## <span style="font-size: 55px;">`explain_lmer(model, details)`</span>

<br>

::: {style="border: 2px solid black; padding: 10px; background-color: #ececec; font-size: 30px"}
✨ A core component of the package, which provides structured, user-friendly interpretations of linear mixed-effects models fitted using `lme4::lmer()`; supports three levels of interpretation via the `details` argument.
:::

<br>

To use this function, first do:
```{r, message=FALSE, echo=TRUE}
model <- lme4::lmer(math ~ cltype + math_old + (cltype | school_id), data = star)
```

## <span style="font-size: 55px;">"general"</span>

::: {style="font-size: 30px"}
Default for `details` argument, which provides a high-level summary of the model, describing its structure, fixed effects, and random effects in a concise yet informative manner:
:::

<div style="margin-top: 20px;"></div>

```{r, echo=TRUE}
explain_lmer(model)
```

## <span style="font-size: 55px;">"fixed"</span>

::: {style="font-size: 30px"}
A detailed explanation of the fixed effects, including interpretations of individual estimates and their implications within the model context:
:::

<div style="margin-top: 20px;"></div>

```{r, echo=TRUE}
fix <- explain_lmer(model, "fixed")
```

---

```{r, echo=TRUE}
attr(fix, "fixed_details")
```

::: callout-tip
## Note

If interaction is fitted to the model, the function is also able to detect and adjust the interpretation based on the interaction structure. 
:::

## <span style="font-size: 55px;">"random"</span>

::: {style="font-size: 30px"}
A detailed explanation of random intercepts, slopes, and variance components to highlight across-group variability:
:::

<div style="margin-top: 20px;"></div>

```{r, echo=TRUE, message=FALSE}
# Nested-grouping example
model_nest <- lmer(math ~ cltype + math_old + (1|system_id) + (1|system_id:school_id), data = star)
random_nest <- explain_lmer(model_nest, "random")
```

---

```{r, echo=TRUE}
attr(random_nest, "var_details")
```

::: callout-tip
## Note

The "random" part is also capable of detecting uncorrelated random effects and multiple (>= 3) random terms.
:::


## (More functions to be added ...)



# Future Work

## (TBD)

## Explore `lme4u`

(Embed final vignettes / website)






